Seasonality in governing upper-ocean dynamics at submesoscales

### 2016-05-11 writeup/rocha_etal.tex
Fixes typos

\section{The LLC2160 model output}
We use the output of a latitude-longitude polar cap (LLC)
MITgcm  numerical simulation  \citep{forget_etal2015}. The output analyzed here, the LLC2160,
is the 1/24$^\circ$ --- 90-vertical-levels simulation from a hierarchy of global
forward numerical solutions on a LLC grid that were
spun up from a ECCO2 adjoint-method state estimate. The LLC2160 was forced by
tides and 6-hourly surface fluxes similarly to the LLC4320 simulation
used by \citet{rocha_etal2016} --- see their appendix D. We chose the second-higher
resolution  simulation (LLC2160) of the LLC hierarchy because it was run for longer:
the LLC2160 output spans two years form March 2011 to April 2013. The model time-step
was 45s and the model variables were saved hourly. LLC hierarchy setup
and control files are available online
(at \texttt{http://mitgcm.org/viewvc/MITgcm/MITgcm$\_$contrib/
llc$\_$hires}). For the LLC hierarchy, the sea-surface height is calculated
using a linearized equation. While the nominal horizontal resolution  of the
LLC2160 simulation is 1/24$^\circ$,
 horizontal wavenumber spectra suggests an effective resolution of about 20 km.
 (see figure \ref{fig3}).

A key aspect of the simulation analyzed here is that it is forced by
the 16 most significant tidal components.
Barotropic tides interact with topography and generate internal
tides that project on mesoscales to submesoscales
\citep{rocha_etal2016}. Hence, tidal forcing fundamentally distinguishes our analysis
from recent modeling studies that investigated upper dynamics
and its seasonality \citep{sasaki_etal2014,qiu_etal2014}.

The region of focus is the northwest Pacific, in the vicinities of the Kuroshio
Extension, where previous studies have
shown strong seasonality in the quasi-gesotrophic surface velocity field.
We analyze a sub-domain of the LLC2160 simulation of about 2000 km$^2$
spanning 155-175$^\circ$E; 25-40$^\circ$N (figure \ref{fig1}a-b).

@@

Focusing on the submesoscale variability, we present estimates of bulk quantities
associated with the the horizontal velocity lateral gradient.
The lateral gradients are calculated using a centered
second-order finite differences scheme. We then calculate
the vertical vorticity $\zeta \equiv v_x - u_y$ and
horizontal divergence $\delta \equiv u_x + v_y$.
These quantities highlight the submesoscale structures in the flow
and hint on their dynamics.

Figures \ref{fig1}a-b show snapshots of vertical vorticity $\zeta$ in early Spring
(April 15) and Fall (October 15). Clearly, a strong seasonality in vorticity
exists: large values of fine-grained vertical vorticity are
observed in early Spring with maximum values as high as $0.9f$, where $f$ is
the local planetary vorticity, and root-mean-square (RMS) of about 0.25. In early
Fall, the situation is the opposite: the vertical vorticity is coarse-grained,
and its local maximum and RMS is no larger than $0.3f$ and $0.1f$, respectively.
Indeed, the seasonal cycle in vorticity/strain is very strong (Figure~\ref{fig2}):
the RMS vertical vorticity is about twice as large in late Winter/early Spring
than it is in late Summer/early Fall. This picture is consistent with
recent studies: shallow baroclinic instabilities energize the submesoscales
in late Winter, drawing from the available potential energy stored in deep mixed
layers \citep{sasaki_etal2014,callies_etal2015,callies_etal2016}.

The bulk of the vertical vorticity is associated with subinertial flows
($T_f = 2\pi/f_0\approx 23.5$h, where $f_0$ is the inertial frequency at
the mean latitude): daily-averaging the velocity fields reduces the vertical
vorticity by less than 20$\%$ and the seasonal cycle remains strong (see
green lines in figure \ref{fig2}). Most of this seasonal cycle is associated
with submesoscale flows: smoothing the velocity field
with a cut-off scale of 100 km reduces the RMS vorticity by at least 40$\%$,
yielding bulk statistics of velocity gradient roughly consistent with AVISO
geostrophic velocities (see lines blue and black in figure \ref{fig2}).

The horizontal divergent part of the flow is significant: the RMS is $\delta/f > 0.1$).
Most of these horizontal divergent flows are supra-inertial and project
on scales smaller than 100 km: daily-averaging or smoothing the velocity fields
 (100 km cutoff) reduces the horizontal divergence by about 70$\%$.
Furthermore, the seasonal cycle in horizontal divergence is out-of-phase with the
seasonal cycle of vertical vorticity: the horizontal divergence is about 30$\%$ larger in late
Summer/early Fall than it is in late Winter/early Spring. The divergence associated
with submesoscale subinertial eddies is small and its seasonal cycle is in phase
with that of vorticity (see green lines).  The seasonal cycle of the horizontal
divergence is roughly in phase with the seasonal cycle of the near-surface (0-50m)
stratification.
\begin{figure*}[ht]
\begin{center}
\includegraphics[width=1.\textwidth]{figs/fig1.pdf}
 \caption{Snapshots of surface vorticity (a and b) and a transect
          of potential density at 165$^\circ$E (c and d). The snapshots were
          taken at 00:00 UTC.}
\vspace{-1.5cm}
 \label{fig1}
 \end{center}
 \end{figure*}

@@

To better understand the projection of these flows onto different horizontal
scales, we calculate wavenumber spectra of kinetic energy and sea-surface height
(SSH) variance. For simplicity we calculate zonal wavenumber spectra at various
latitudes. The two-year time averaged and spatial linear trend were
removed before calculating discrete Fourier transforms.Statistical significance
is obtained by averaging over latitude and time. Using conservative temporal and
spatial decorrelation estimate of 10 days and 250 km, we obtain roughly 50
independent realizations of the spectra.

Figure \ref{fig3}a depicts the horizontal wavenumber spectra of surface KE in
April and October. At scales larger than 20km, the spectra based on hourly
velocity snapshots are nearly indistinguishable within 95$\%$ confidence level
(compare solid lines in figure \ref{fig3}a). Consistent with the results of
\citet{rocha_etal2016} who analyzed the LLC4320 output in Drake Passage,
there is significant high-frequency variabilty at submesoscales. Daily-averaging
the velocity field suppresses spatial variability at scales small than about 250
km both in April and October (compare solid lines against dashed lines in figure
\ref{fig3}a). But this suppression is more dramatic in October, when the horizontally
divergent flows peak. At scales
smaller than 100 km, 40$\%$ of the surface KE in April a is accounted for by suprainertial
flows as opposed to 77$\%$ in October. The seasonality of subinertial submesoscale flows
is significant, consistent with the results of \citet{sasaki_etal2014}, which are
based on snapshots of a model without tidal forcing.

These high-frequency submesoscale flows significantly project on the sea-surface.
In October, when the horizontally divergent flows peak and the submesoscale
balanced flows are minimum, there is a dramatic difference between the spectra
based on hourly and daily-averaged SSH: at scales smaller than 100 km, the spectra
of hourly SSH roughly follows a $-2$ power-law, whereas the spectra of daily-averaged
SSH roughly follows a $-5$ power-law. At scales
smaller than 100 km, 44$\%$ of the surface KE in April is accounted for by suprainertial
flows as opposed to 86$\%$ in October! The out-of-phase seasonal cycle of subinertial
and suprainertial flows conspire to yield no significant seasonality in the spectra
of hourly SSH.

@@

The surface KE spectra of subinertial flows very roughly follow a $-2$ power-law.
The high-frequency flows are split near-inertial flows, tidal signals and their
higher harmonics. Hence these high-frequency flows are dominated by inertia-gravity
waves. (Note that the inertial peak nearly merges with the
diurnal tide peak.) There is statistically significant seasonality. The seasonal
cycle of the sub-inertial motions (at least for periods between 5-20 days)
is out-of-phase with the seasonal cycle of supra-inertial flows: the former
are more energetic in April and the latter in October.

@@

That the surface velocity and SSH submesoscale (10-100 km) variability may be
dominated by ageostrophic flows in
Summer/Fall has consequences for the interpretation of data from
the future SWOT and COMPIRA altimeter missions,
which will deliver SSH measurements at submesoscales. To the extent that
high-frequency flows are dominated by spatially incoherent internal tides and other
internal waves, it may be very difficult (if not impossible) to
separate submesoscale SSH variability associated with geostrophic motions
from high-frequency, ageostrophic flows. Thus, previous claims that
one will be able to easily obtain submesoscale surface
geostrophic velocities and monitor such seasonal cycle on global scales
\citep{sasaki_etal2014,qiu_etal2014} are overstated.

>>>>

\section{The LLC2160 model output}
We use the output of a latitude-longitude polar cap (LLC)
MITgcm  numerical simulation  \citep{forget_etal2015}. The output analyzed here, the LLC2160,
is the 1/24$^\circ$ --- 90-vertical-levels simulation from a hierarchy of global
forward numerical solutions on a LLC grid that were
spun up from a ECCO2 adjoint-method state estimate. The LLC2160 was forced by
tides and 6-hourly surface fluxes similarly to the LLC4320 simulation
used by \citet{rocha_etal2016} --- see their appendix D. We chose the second-higher
resolution  simulation (LLC2160) of the LLC hierarchy because it was run for longer:
the LLC2160 output spans two years from March 2011 to April 2013. The model time-step
was 45s and the model variables were saved hourly. LLC hierarchy setup
and control files are available online
(at \texttt{http://mitgcm.org/viewvc/MITgcm/MITgcm$\_$contrib/
llc$\_$hires}). For the LLC hierarchy, the sea-surface height is calculated
using a linearized equation. While the nominal horizontal resolution  of the
LLC2160 simulation is 1/24$^\circ$,
 horizontal wavenumber spectra suggests an effective resolution of about 20 km.
 (see figure \ref{fig3}).

A key aspect of the simulation analyzed here is that it is forced by
the 16 most significant tidal components.
Barotropic tides interact with topography and generate internal
tides that project on mesoscales to submesoscales
\citep{rocha_etal2016}. Hence, tidal forcing fundamentally distinguishes our analysis
from recent modeling studies that investigated upper-ocean dynamics
and its seasonality \citep{sasaki_etal2014,qiu_etal2014}.

The region of focus is the northwest Pacific, in the vicinities of the Kuroshio
Extension, where previous studies have
shown strong seasonality in submesoscale turbulence.
We analyze a sub-domain of the LLC2160 simulation of about 2000 km$^2$
spanning 155-175$^\circ$E; 25-40$^\circ$N (figure \ref{fig1}a-b).

@@

Focusing on the submesoscale variability, we present estimates of bulk quantities
associated with the lateral gradient of the horizontal velocity.
The lateral gradients are calculated using a centered
second-order finite differences scheme. We then calculate
the vertical vorticity $\zeta \equiv v_x - u_y$ and
horizontal divergence $\delta \equiv u_x + v_y$.
These quantities highlight the submesoscale structures in the flow
and hint on their dynamics.

Figures \ref{fig1}a-b show snapshots of vertical vorticity $\zeta$ in early Spring
(April 15) and Fall (October 15). Clearly, a strong seasonality in vorticity
exists: large values of fine-grained vertical vorticity are
observed in early Spring with maximum values as high as $0.9f$, where $f$ is
the local planetary vorticity, and root-mean-square (RMS) of about 0.25. In early
Fall, the situation is the opposite: the vertical vorticity is coarse-grained,
and its local maximum and RMS is no larger than $0.3f$ and $0.1f$, respectively.
Indeed, the seasonal cycle in vorticity is very strong (Figure~\ref{fig2}):
the RMS vertical vorticity is about twice as large in late Winter/early Spring
than it is in late Summer/early Fall. This picture is consistent with
recent studies: shallow baroclinic instabilities energize the submesoscales
in late Winter, drawing from the available potential energy stored in deep mixed
layers \citep{sasaki_etal2014,callies_etal2015,callies_etal2016}.

The bulk of the vertical vorticity is associated with subinertial flows
($T_f = 2\pi/f_0\approx 23.5$h, where $f_0$ is the inertial frequency at
the mean latitude): daily-averaging the velocity fields reduces the vertical
vorticity by less than 20$\%$ and the seasonal cycle remains strong (see
green lines in figure \ref{fig2}). Most of this seasonal cycle is associated
with submesoscale flows: smoothing the velocity field
with a Hanning filter with cut-off scale of 100 km reduces the RMS vorticity by
at least 40$\%$, yielding bulk statistics of velocity gradient roughly consistent
with AVISO geostrophic velocities (see lines blue and black in figure \ref{fig2}).

The horizontal divergent part of the flow is significant: the RMS is $\delta/f > 0.1$).
Most of these horizontal divergent flows are supra-inertial and project
on scales smaller than 100 km: daily-averaging or smoothing the velocity fields
 (100 km cutoff) reduces the horizontal divergence by about 70$\%$.
Interestingly, the seasonal cycle in horizontal divergence is out-of-phase with the
seasonal cycle of vertical vorticity: the horizontal divergence is about 30$\%$ larger in late
Summer/early Fall than it is in late Winter/early Spring. The divergence associated
with submesoscale subinertial eddies is small and its seasonal cycle is in phase
with that of vorticity (see green lines).  The seasonal cycle of the horizontal
divergence is roughly in phase with the seasonal cycle of the near-surface (0-50m)
stratification.
\begin{figure*}[ht]
\begin{center}
\includegraphics[width=1.\textwidth]{figs/fig1.pdf}
 \caption{Snapshots of surface vorticity (a and b) and a transect
          of potential density at 165$^\circ$E (c and d). The snapshots were
          taken at 00:00 UTC.}
\vspace{-1.5cm}
 \label{fig1}
 \end{center}
 \end{figure*}

@@

To better understand the projection of these flows onto different horizontal
scales, we calculate wavenumber spectra of kinetic energy and sea-surface height
(SSH) variance. For simplicity we calculate zonal wavenumber spectra at various
latitudes. The two-year time averaged and spatial linear trend were
removed, and the ensuing fields were Hanning spectral ``window''. Statistical significance
is obtained by averaging over latitude and time. Using conservative temporal and
spatial decorrelation estimate of 10 days and 250 km, we obtain roughly 50
independent realizations of the spectra.

Figure \ref{fig3}a depicts the horizontal wavenumber spectra of surface KE in
April and October. At scales larger than 20km, the spectra based on hourly
velocity snapshots are nearly indistinguishable within 95$\%$ confidence level
(compare solid lines in figure \ref{fig3}a). Consistent with the results of
\citet{rocha_etal2016} who analyzed the LLC4320 output in Drake Passage,
there is significant high-frequency variabilty at submesoscales. Daily-averaging
the velocity field suppresses spatial variability at scales smaller than about 250
km both in April and October (compare solid lines against dashed lines in figure
\ref{fig3}a). But this suppression is more dramatic in October, when the horizontally
divergent flows peak. At scales
smaller than 100 km, 40$\%$ of the surface KE in April is accounted for by supra-inertial
flows as opposed to 77$\%$ in October. The seasonality of subinertial submesoscale flows
is significant, consistent with the results of \citet{sasaki_etal2014}, which are
based on snapshots of a model without tidal forcing.

These high-frequency submesoscale flows significantly project on the sea-surface.
In October, when the horizontally divergent flows peak and the submesoscale
balanced flows are minimum, there is a dramatic difference between the spectra
based on hourly and daily-averaged SSH: at scales smaller than 100 km, the spectra
of hourly SSH roughly follows a $-2$ power-law, whereas the spectra of daily-averaged
SSH roughly follows a $-5$ power-law. At scales
smaller than 100 km, 44$\%$ of the surface KE in April is accounted for by suprainertial
flows as opposed to 86$\%$ in October! The out-of-phase seasonal cycle of subinertial
and suprainertial flows conspire to yield nearly insignificant seasonality in the spectra
of hourly SSH.

@@

The surface KE spectra of subinertial flows very roughly follow a $-2$ power-law.
The high-frequency flows are split near-inertial flows, tidal signals and their
higher harmonics. Hence these high-frequency flows are dominated by inertia-gravity
waves. (Note that the inertial peak nearly merges with the
diurnal tide peak.) There is statistically significant seasonality. The seasonal
cycle of the sub-inertial motions (at least for periods between 5-20 days)
is out-of-phase with the seasonal cycle of supra-inertial flows: the former flows
are more energetic in April and the latter in October.

@@

That the  surface velocity and SSH submesoscale variability may be
dominated by ageostrophic flows in
Summer/Fall has consequences for the interpretation of data from
the future SWOT and COMPIRA altimeter missions,
which will deliver SSH measurements at submesoscales. To the extent that
high-frequency flows are dominated by spatially incoherent internal tides and other
internal waves, it may be very difficult (if not impossible) to
separate SSH submesoscale variability associated with geostrophic motions
from high-frequency, ageostrophic flows. Thus, previous claims that
one will be able to easily obtain submesoscale surface
geostrophic velocities and monitor such seasonal cycle on global scales
\citep{sasaki_etal2014,qiu_etal2014} are overstated.
